A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-02-25 pkg/colorspace/paper/Toolbox/article.Rmd
further streamlining
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@455 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

```{r eustockmarkets-plot, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
q4 <- qualitative_hcl(4)
par(mar = c(5, 4, 1, 1))
plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
```

@@

```{r titanic-plot, echo = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
par(mar = c(5, 4, 1, 1))
spineplot(ttnc, col = sequential_hcl(2, "Purples 3"))
```

@@

```{r iris-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")
```

@@

```{r diamonds-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
```

@@

```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, out.width = "\\textwidth", fig.cap = "Foo."}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
  "Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
  "Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
  "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
  nrow = 5
)
```

@@

The illustrations below show how basic qualitative, sequential, and diverging
palettes are constructed in HCL space (the corresponding mathematical equations
are provided in the [construction details](hcl_palettes.html#construction-details)).
In the qualitative "Set 2" palette below, the
hue traverses the entire color "wheel" (from 0 to 360 degrees) while keeping
chroma and luminance (almost) constant (C = 60 and L = 70).

```{r spec-qualitative, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(qualitative_hcl(100, "Set 2"))
```

Note that due to the restrictions of the HCL color space, some of the green/blue
colors have a slightly smaller maximum chroma resulting in a small dip in the chroma
curve. This is fixed automatically (by default) and is hardly noticable in visualizations,
though.

The sequential "Blues 2" palette below employs a single hue (H = 260) and a monotonically
increasing luminance trajectory (from dark to light). Chroma simply decreases monotonically
along with increasing luminance.

```{r spec-sequential, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(sequential_hcl(100, "Blues 2"))
```

Finally, the diverging "Blue-Red" palette is depicted below. It simply combines a blue
and a red sequential single-hue palette (similar to the "Blues 2" palette discussed above).
Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are balanced between both arms.
In the center the palette passes through a light gray (with zero chroma) as the neutral
value.

```{r spec-diverging, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(diverging_hcl(100, "Blue-Red"))
```

@@

```{r spec-rainbow, fig.width = 5, fig.height = 7.5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(rainbow(100), rgb = TRUE)
```

>>>>

@@

```{r eustockmarkets-titanic, echo = FALSE, fig.height = 4, fig.width = 10, out.width = "\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
par(mar = c(5, 4, 1, 1), mfrow = c(1, 2))
plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
spineplot(ttnc, col = sequential_hcl(2, "Purples 3"))
```

@@

```{r iris-ggplot, eval = FALSE}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")
```

@@

```{r diamonds-ggplot, eval = FALSE}
dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
```

```{r iris-diamonds, echo = FALSE, fig.height = 4, fig.width = 5, out.width = "0.49\\textwidth", fig.cap = "Foo."}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")
set.seed(1)
dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
```

@@

```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, out.width = "\\textwidth", fig.cap = "Foo."}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
  "Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
  "Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
  "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
  nrow = 5, line = 5
)
```

@@

FIXME: Point to previous plots

Finally, the diverging "Green-Brown" palette is depicted below. It simply combines a green
and a brown/yellow sequential single-hue palette, both with triangular chroma trajectory.
Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are balanced between both arms.
In the center the palette passes through a light gray (with zero chroma) as the neutral
value.

@@

```{r specplot, fig.width = 5, fig.height = 7, out.width = "0.49\\textwidth", fig.cap = "Foo."}
specplot(diverging_hcl(100, "Green-Brown"), rgb = TRUE)
specplot(rainbow(100), rgb = TRUE)
```
